p-adic stability in linear algebra

### 2015-01-05 examples/stability.tex
Section on LU factorization compactified

@@

\begin{abstract}
Using the differential precision methods developed previously by the same authors,
we study the $p$-adic stability of standard operations on matrices and vector 
spaces. We demonstrate that lattice-based methods surpass naive methods in many
applications, such as matrix multiplication and sums and intersections of subspaces.
We also analyze determinants and characteristic polynomials using these differential methods.
We supplement these observations with numerical experiments.
\end{abstract}

@@

\noindent
{\bf About the choice of the function $h$}.
At the beginning of \S \ref{ssec:boundLambdaf}, we have introduced a 
function $h$ in our differential equation. The aim of this was to be 
more flexible and possibly get this way better bounds on $\Lambda(f)$. 
In this paragraph, we discuss a recommended choice for it.

Actually, we do recommend taking $h$ as the identity function 
\emph{but} changing the norm on the codomain of $h$ at the same time.
More precisely, if $F$ is equipped with the norm $\Vert \cdot \Vert_F$, 
we recommand endowing it with the second norm defined by
$\Vert x \Vert'_F = \lambda \cdot \Vert x \Vert_F$ ($x \in F)$
where $\lambda$ is a positive real number that we shall choose later 
and taking $h : (F, \Vert \cdot \Vert_F) \to (F, \Vert \cdot \Vert'_F)$ 
acting as the identity on the underlying vector spaces. Doing this, the 
function $\Lambda(h)$ maps $x$ to $x + \log \lambda$ ($x \in \R$) and we 
choose $\Lambda_h = \Lambda(h)$.

We believe that a good way to optimize $\lambda$ is to choose it so that 
the two quantities in the minima in Proposition \ref{prop:boundLambdaf2} 
agree. With the expression we got just above for $\Lambda_h$, our 
heuristic yields $\log \lambda = \mu$, \emph{i.e.} $\lambda = e^\mu$.
Making this choice, the norms of $f(x)$ and $h(x)$ are comparable --- 
at least, on the domain we are interested in --- and we do not loose so 
much when we write that the norm of the couple $(f(x), h(x))$ is the 
maximum of those of $f(x)$ and $h(x)$.

@@

\noindent
{\bf Smith form.}
In what follows, we will often use the \emph{Smith decomposition} of a 
matrix $M \in M_{n,m}(K)$: we recall that it is a factorization of $M$ 
as a product $U_M \cdot \Delta_M \cdot V_M$ when $U_M$ and $V_M$ are 
unimodular (square) matrices of size $n$ and $m$ respectively, with 
coefficients in $\O_K$ and $\Delta_M$ satisfies the two following
conditions:

\noindent
(i) its $(i,j)$-th entry vanishes as soon as $i \neq j$

\noindent
(ii) its $(i,i)$-th entry divides its $(j,j)$-th entry when $i
\leq j$.

\noindent
Such a factorization is \emph{not} unique but the valuations of
the ``diagonal'' entries of $\Delta_M$ are. Smith decompositions
can be easily computed by performing elementary operations on rows and columns.
Regarding to precision, one can prove that if the entries of $M$ are all 
known up to some $O(\pi^N)$, it is possible to compute a factorization 
$U_M \cdot \Delta_M \cdot V_M$ where $U_M$ and $V_M$ are \emph{exact}
matrices satisfying the above conditions and where $\Delta_M$ is known
at precision $O(\pi^N)$ and satisfies~(i) and~(ii) modulo $\pi^N$.

@@

In terms of precision, Proposition \ref{prop:detmatrix} implies that if 
$M$ is given at flat precision $O(\pi^N)$ with $N > v$, then $\det(M)$ 
is known at precision $O(\pi^{N+v})$, so we are gaining $v$ significant 
digits in absolute precision\footnote{We note that the valuation of 
$\det(M)$ is $v + m_n$; thus if we are considering \emph{relative} 
precisions then we should say that we are loosing $m_n$ significant 
digits.}.
Furthermore, there exists a easy method for computing the determinant
of $M$ with this optimal precision. It consists in first computing an
approximate Smith decomposition of $M$:
$$M = \tilde U_M \cdot \tilde \Delta_M \cdot \tilde V_M.$$
Here $\tilde U_M$ and $\tilde V_M$ are exact unimodular matrices and
$\tilde \Delta_M$ is known at precision $O(\pi^N)$, is diagonal modulo
$\pi^N$ and its diagonal entries $a_1, \ldots, a_m$ have valuations 
$m_1, \ldots, m_n$ respectively.
We then have $\det(M) = \det(\tilde \Delta_M)$ and factoring $a_i$ on
the $i$-th row of $\tilde \Delta_M$, we obtain
$\det(\tilde \Delta_M) = a_1 \cdots a_m \cdot U$
where $U$ is a matrix which is congruent to the identity modulo 
$\pi^{N-m_n}$. Therefore we get:
$$\det(M) \equiv a_1 \cdots a_m \pmod{\pi^{N+v}}$$
from what we see that we can compute $\det(M)$ at the required precision
by simply evaluating the product $a_1 \cdots a_m$.

@@

\todo{Improve this Subsection.}

In this Subsection, we apply our framework to the case of $LU$ factorization. The computation of the differential of this factorization has already been done in \cite[Appendix B]{caruso-roe-vaccon:14a}. It remains to determine under which condition Proposition \ref{prop:precision} applies.

\begin{deftn}
Let $n \in \mathbb{Z}_{\geq 0}$ and  $O_n \subset K^{n \times n}$ be the set of the matrices such that all principal minors are non-vanishing. Let $U_n$ (resp. $L_n$)  be the $K$-vector spaces of the upper-triangular (resp. the strictly lower-triangular) matrices in $K^{n \times n}.$ Let $L_n^u \subset K^{n \times n} $ be the set of lower-triangular matrices with $1$'s one the diagonal. We define $D : O_n \rightarrow L_n^u \times U_n$ to be the mapping sending a matrix $M$ to $(L,M)$ such that $LU$ is the LU factorization of $M.$
\end{deftn}

This example has been previously studied by X.Caruso in \cite{caruso:12a}, but without using differential precision.
We show here that the framework of Subsection \ref{ssec:boundLambdaf} applies directly.

\subsubsection*{Computation of the differential of $D$} 
We begin with some notations:
\begin{deftn}
We denote by $\textrm{low}$ (resp. $\textrm{up}$) the canonical projection $K^{n \times n} \rightarrow L_n$ (resp. $K^{n \times n} \rightarrow U_n$).
\end{deftn}

Then, an explicit formula for $D'$ has been computed in \cite[Appendix B]{caruso-roe-vaccon:14a} :
\begin{prop}
Let $(M,L,U) \in O_n \times L_n^u \times U_n$ be such that $M=LU.$ Then,

\begin{tabular}{cccc}
$D'(M) :$ &  $K^{n \times n}$ & $\rightarrow$ & $L_n \times U_n$ \\ 
 & $\mathrm{d}M$ & $ \mapsto$ & $(L \textrm{low}(L^{*-1} \mathrm{d}M U^{-1}), \textrm{up}(L^{*-1} \mathrm{dM} U^{-1}) U).$ \\ 
\end{tabular} 
\end{prop}

A right inverse of $D'(M)$ can be made explicit:
\begin{rem} \label{rem:rightinverseLU}
Let \begin{tabular}{cccc}
$E(M) :$ & $L_n \times U_n$  & $\rightarrow$ &  $K^{n \times n}$\\ 
 & $(A,B)$ & $ \mapsto$ & $AU+LB$. \\ 
\end{tabular}

Then $D'(M)  \circ E(M) = Id.$
\end{rem}

\subsubsection*{Applying Subsection \ref{ssec:boundLambdaf}}

In order to apply Subsection \ref{ssec:boundLambdaf}, $f$ will be a translation of $D,$ so as to ensure $g(0) \neq 0.$ Namely,  let $(M_0,L_{0},U_{0}) \in O_n \times L_n^u \times U_n$ be such that $M_0 = L_{0}U_{0},$ we define

\begin{tabular}{cccc}
$f :$ &  $K^{n \times n}$ & $\rightarrow$ & $L_n \times U_n$ \\ 
 & $M$ & $ \mapsto$ & $(L_{M+M_0}-L_{0}, U_{M+M_0}-U_{0}).$ \\ 
\end{tabular}

Then, $f(0)=0$ and \[f'(M) \cdot \mathrm{d}M = (L_{M+M_0} \textrm{low}(L_{M+M_0}^{-1} \mathrm{d}M U_{M+M_0}^{-1}), \textrm{up}(L_{M+M_0}^{-1} \mathrm{dM} U_{M+M_0}^{-1}) U).\]

We now proceed with the three steps of the framework of Subsection \ref{ssec:boundLambdaf}.

\textbf{The $C$ constant:}

\begin{lem}
For $C = \max (\Vert U_{0} \Vert,\Vert L_{0} \Vert ),$ $B_{K^{n \times n}}(1) \subset f'(0) \cdot B_{K^{n \times n}}(C).$ 
\end{lem}
\begin{proof}
Apply Remark \ref{rem:rightinverseLU}.
\end{proof}

\textbf{The differential equation:}

\begin{lem} Let $g=g_{M_0}$ be defined by:
\begin{align*}
g(A,B) \cdot X =& ( (L_0^*+A) \textrm{low} \left( (L_0^*+A)^{-1}X(U_0+B)^{-1} \right), \\
& \: \textrm{up} \left( (L_0^*+A)^{-1}X(U_0+B)^{-1} \right) (U_0+B) ).\\
\end{align*}
Then, \[f' = g \circ f.\]
\end{lem}

\textbf{Determining $\mu$ and $\nu$}

Let $\vert \: \vert $ be an ultrametric norm on $K$ and $\Vert \: \Vert$ be the corresponding subordinate matrix norm. To determine $\mu$ and $\nu,$ we first use the following lemma

\begin{lem}
Let $(U,V) \in K^{n \times n} \times K^{n \times n}.$ If $\Vert V \Vert < \Vert U \Vert$ then $\Vert U+V \Vert = \Vert U \Vert.$

If $U \in GL_n(K)$ and $\Vert U \Vert < \Vert U^{-1} \Vert^{-1},$ then $\Vert (U+V)^{-1} \Vert = \Vert U^{-1} \Vert.$

Finally $\Vert U^{-1} \Vert^{-1} \leq \Vert U \Vert.$
\end{lem}

This leads to the following upper-bound on the norm of $g(A,B)$.

\begin{prop}
Let $(A,B) \in L_n \times U_n.$ be such that $\Vert A \Vert < \Vert L_0^{*-1} \Vert $ and $\Vert B \Vert <\Vert U_0^{*-1} \Vert. $ Then 
\[ \Vert g(A,B) \cdot X \Vert \leq \max \left( \Vert L_0^* \Vert \Vert L_0^{*-1} \Vert  \Vert U_0^{*-1} \Vert, \Vert U_0^* \Vert \Vert U_0^{*-1} \Vert  \Vert L_0^{*-1} \Vert \right) \times \Vert X \Vert . \]
\end{prop}

\begin{prop}
For $f$ and $g$ as before, we can take $\nu = \min ( \Vert L_0^{-1} \Vert ^{-1} ,\Vert U_0^{-1} \Vert ^{-1} ),$ and \[ \mu = \max \left( \Vert L_0^* \Vert \Vert L_0^{*-1} \Vert  \Vert U_0^{*-1} \Vert, \Vert U_0^* \Vert \Vert U_0^{*-1} \Vert  \Vert L_0^{*-1} \Vert \right).\]
\end{prop}

\textbf{Conclusion}

Consequently, we can apply Proposition \ref{prop:boundLambdaf2} and Proposition 3.12 of \cite{caruso-roe-vaccon:14a}:

\begin{prop}
For any $\rho \in (0,1],$ if $r<e^{ \nu -2(\alpha +\mu)+\log \left( \frac{\rho}{C} \right)},$ the equation \ref{eq:firstorder} for $f$ is satisfied.
\end{prop}

\begin{ex}
If $\Vert L_0\Vert = \Vert L_0^{-1}\Vert =\Vert U_0\Vert =  \Vert U_0^{-1}\Vert = \rho=1$ and $p=2,$ then Proposition \ref{prop:precision} can be applied with $r=2^{-4}.$
\end{ex}

\subsubsection*{Numerical experiments}

\todo{Add numerical experiments...}

>>>>

\newcommand{\low}{\text{\rm low}}
\newcommand{\up}{\text{\rm up}}

@@

\begin{abstract}
Using the differential precision methods developed previously by the same authors,
we study the $p$-adic stability of standard operations on matrices and vector 
spaces. We demonstrate that lattice-based methods surpass naive methods in many
applications, such as matrix multiplication and sums and intersections of subspaces.
We also analyze determinants, characteristic polynomials and LU factorization using these differential methods.
We supplement our observations with numerical experiments.
\end{abstract}

@@

\begin{rem}
In certain situations,
it may happen that the function $f$ is solution of a simplier 
differential equation of the form $f' = g \circ f$. If this holds, 
Proposition \ref{prop:boundLambdaf2} gives the bound $\Lambda(f)_{\leq 
2} (x) \leq 2(x + \alpha + \mu) - \nu$ for $x \leq \nu - \mu - \alpha$.

Beyond this particular case, our general recommandation for choosing the 
function $h$ is to endow $F$ with the second norm $\Vert x \Vert'_F = 
e^\mu \cdot \Vert x \Vert_F$ ($x \in F$) and to take $h : (F, \Vert 
\cdot \Vert_F) \to (F, \Vert \cdot \Vert'_F)$ acting as identity on the
underlying vector spaces. The function $\Lambda(h)$ the maps $x$ to $x + 
\mu$ ($x \in \R$) and we can choose $\Lambda_h = \Lambda(h)$.
\end{rem}

@@

@@

In terms of precision, Proposition \ref{prop:detmatrix} implies that if 
$M$ is given at flat precision $O(\pi^N)$ with $N > v$, then $\det(M)$ 
is known at precision $O(\pi^{N+v})$, so we are gaining $v$ significant 
digits in absolute precision\footnote{We note that the valuation of 
$\det(M)$ is $v + m_n$; thus if we are considering \emph{relative} 
precisions then we should say that we are loosing $m_n$ significant 
digits.}.
Furthermore, there exists a easy method for computing the determinant
of $M$ with this optimal precision. It consists in first computing an
approximate Smith decomposition of $M$, that is a writing
$M = \tilde U_M \cdot \tilde \Delta_M \cdot \tilde V_M$
where $\tilde U_M$ and $\tilde V_M$ are exact unimodular matrices and
$\tilde \Delta_M$ is known at precision $O(\pi^N)$, is diagonal modulo
$\pi^N$ and has diagonal entries $a_1, \ldots, a_m$ of valuations 
$m_1, \ldots, m_n$ respectively.
We then have $\det(M) = \det(\tilde \Delta_M)$ and factoring $a_i$ on
the $i$-th row of $\tilde \Delta_M$, we obtain
$\det(\tilde \Delta_M) = a_1 \cdots a_m \cdot U$
where $U$ is a matrix which is congruent to the identity modulo 
$\pi^{N-m_n}$. Therefore we get:
$$\det(M) \equiv a_1 \cdots a_m \pmod{\pi^{N+v}}$$
from what we see that we can compute $\det(M)$ at the required precision
by simply evaluating the product $a_1 \cdots a_m$.

@@

In this subsection, we study the LU factorization.
%This example has been previously studied by Caruso in \cite{caruso:12a}, 
%but without using differential precision. We are going to show here that 
%the framework of \S \ref{ssec:boundLambdaf} applies directly.
We first introduce several notations.
We denote by $\Vert \cdot \Vert$ the subordinate matrix norm on $M_n(K)$
and, given a positive real number $C$, we let $B(C)$ be the closed ball 
in $M_n(K)$ centered at the origin of radius $C$. We consider in addition
the following subsets of $M_n(K)$:

\noindent $\bullet$
$O_n$ is the open subset of matrices whose all principal 
minors do not vanish (we recall that the latest condition implies the
existence and the unicity of a LU factorization);

\noindent $\bullet$
$U_n$ is the sub-vector space of upper-triangular 
matrices;

\noindent $\bullet$
$L_n$ (resp. $L_n^u$) is the sub-affine space space of 
nilpotent (resp. unipotent) lower-triangular matrices.

\noindent
{\bf Calculus and precision.}
We choose to normalize the LU factorization by requiring that $L$ is 
unipotent and denote by $D : O_n \to L_n^u \times U_n$ the function 
modeling this decomposition. The computation of the differential of $D$ 
has already been done in \cite[Appendix B]{caruso-roe-vaccon:14a}. For
$M \in \O_n$ with $D(M) = (L,U)$, the linear mapping $D'(M)$ is given 
by:
$$\begin{array}{l}
dM \mapsto (L \cdot \low(dX), \up(dX) \cdot U) \smallskip \\
\hspace{1cm}\text{with } dX = L^{-1} \cdot dM \cdot U^{-1}
\end{array}$$
where $\low$ (resp. $\up$) denotes the canonical projection of $M_n(K)$ 
onto $L_n$ (resp. $U_n$). 
It is easily checked that $D'(M)$ is bijective with inverse given by
$(A,B) \mapsto AU+LB$.

We now want to apply Proposition \ref{prop:boundLambdaf2}
in order to derive a concrete result on precision. We pick $M_0 \in
O_n$ and write $D(M_0) = (L_0,U_0)$. We consider the ``translated'' 
function $f$ taking $M$ to $D(M_0+M)-D(M_0)$. We then have $f(0) = 0$ 
and $f'(M) = D'(M_0+M)$. Using the explicit description of the inverse
of $D'(M_0+M)$, we find $B(1) \subset f'(0) \cdot B(C)$
for $C = \max(\Vert U_0 \Vert, \Vert L_0 \Vert)$. Moreover $f$ satisfies 
the differential equation $f' = g \circ f$ where $g$ is defined by:
\begin{equation}
\label{eq:gAB}
\begin{array}{l}
g(A,B)(X) = \big( (L_0+A) \cdot \low(Y), \up(Y) \cdot (U_0+B) \big)
\smallskip \\
\hspace{1.8cm}\text{with } Y = (L_0+A)^{-1} \cdot X \cdot (U_0+B)^{-1}.
\end{array}
\end{equation}
Let $\kappa(S) = \Vert S \Vert \cdot \Vert S^{-1} \Vert$ denote the 
condition number of a matrix $S$. Remarking that $\Vert S + T 
\Vert = \Vert S \Vert$ if $\Vert T \Vert < \Vert S \Vert$ and $\Vert (S 
+ T)^{-1} \Vert = \Vert S^{-1} \Vert$ if $\Vert T \Vert < \Vert S^{-1} 
\Vert^{-1}$, we deduce from \eqref{eq:gAB} that
$\Vert g(A,B) \Vert \leq \max 
\big( \kappa(L_0) \Vert U_0^{-1} \Vert, 
\kappa(U_0) \Vert L_0^{-1} \Vert \big)$
provided that $\Vert A \Vert < \Vert L_0^{-1} \Vert^{-1}$ and 
$\Vert B \Vert < \Vert U_0^{-1} \Vert^{-1}$.
Combining this with Proposition \ref{prop:boundLambdaf2} and 
\cite[Proposition.~3.12]{caruso-roe-vaccon:14a}, we finally find that 
Eq.~\eqref{eq:firstorder} holds as soon as 
$$\begin{array}{l}
\displaystyle \frac \rho r > \Vert p \Vert^{-\frac{2p}{p-1}} \cdot
\max ( \Vert L_0 \Vert, \Vert U_0 \Vert) \cdot
\max ( \Vert L_0 ^{-1} \Vert, \Vert U_0 ^{-1} \Vert) \cdot{} \\
\hspace{3.5cm} \max \big( \kappa(L_0) \Vert U_0^{-1} \Vert, 
  \kappa(U_0) \Vert L_0^{-1} \Vert \big)^2.
\end{array}$$

\noindent
{\bf Numerical experiments.}
\todo{Do it.}
